Appendix 2: simulating species interactions

### 2015-11-05 Harris_2015/manuscript-materials/draft.tex
add fakedata and updates

\begin{spacing}{1.9}
\begin{flushleft}
\chead{\ifthenelse{\value{page}=1}{}{\textit{Species interactions in Markov networks}}}
\renewcommand{\headrulewidth}{0pt}

@@

I then generated landscapes with up to 20 interacting species at 25,
200, or 1600 sites using three increasingly complex models (50
replicates for each combination of size and model). As described in
Appendix {[}{[}X{]}{]}, I randomly drew the ``true'' coefficient values
for each replicate landscape so that most species pairs interacted
negligibly, a few pairs interacted very strongly, and competition was
three times more common than facilitation.

The first set of 20-species landscapes, like the landscapes with three
species, were generated directly from a Markov network to ensure that
the model could recover the parameters used to generate the ``observed''
co-occurrence data. Then, I added two environmental factors that varied
from location to location across the simulated landscapes, and simulated
a new set of co-occurrence data so that species' \(\alpha\) coefficients
depended on the local environment. The latter set of simulated
landscapes provide an important test of the methods' ability to
distinguish co-occurrence patterns that were generated from pairwise
biotic interactions from those that were generated by external forces
like abiotic environmental filtering. This task was made especially
difficult because---as with most analyses of presence-absence data for
co-occurrence patterns---the inference procedure did not have access to
any information about the environmental or spatial variables that helped
shape the landscape (cf Connor et al. 2013). I generated the final set
of landscapes by simulating the population dynamics of 20 interacting
species, as described in Appendix {[}{[}X{]}{]}.

\noindent \textbf{Recovering species interactions from simulated data.}
I compared seven techniques for determining the sign and strength of the
associations between pairs of species from simulated data (Appendix 3).
First, I used the rosalia package (Harris 2015a) to fit Markov network
models, as described above. For the analyses with 20 species, I added a
weak regularizer (equivalent to a logistic prior with scale 2) to ensure
that the model estimates were always finite {[}{[}Appendix{]}{]}.

@@

\noindent
\textbf{Twenty species.} Despite some variability across contexts
(Figure 3), the four methods that controlled for indirect effects (the
Markov network, the generalized linear models, and the two partial
correlation-based methods) outperformed those that did not. Of these,
the Markov network consistently performed the best. As anticipated by
Lee and Hastie (2012), generalized linear models closely approximated
the Markov network estimates, especially when the data sets were very
large. As reviewed in Gotelli and Ulrich (2009), however, most analyses
in this field involve fewer than 50 sites, so model performance on the
smaller landscapes where the gap was larger is most important. As shown
in Appendix {[}{[}X{]}{]}, the standard errors associated with the
estimates in Figure 3 are small (less than 0.01), so the differences
among methods should not be attributed to sampling error.

@@

Finally, we can evaluate the models' statistical inferences (focusing on
the first two simulation types, for which the true interaction rates are
easiest to interpret). The Markov network's approximate Type I error
rate (defined here as its probability of rejecting the null hypothesis
for a pair of species where \(|\beta_{ij}|<0.1\)) depended on the
simulation type: 0.03 for simulations that matched the model's
assumptions, versus 0.15 for simulations that included environmental
heterogeneity {[}{[}Appendix{]}{]}. For the null model, these rates were
0.31 and 0.50, respectively---far higher than the nominal 0.05 rate.
Figure 4C shows, across a range of true interaction strengths, the
probability that the null model or the Markov network will predict the
wrong sign of the interaction with 95\% confidence. The null model makes
such errors nearly 8 times as often as the Markov network, even though
it only reject the null hypothesis twice as often overall
{[}{[}Appendix{]}{]}. The Markov network's errors were also more
concentrated around 0, so it never misclassified strong interactions
like the null model did.

@@

\noindent \textbf{Figure 1: A.} A small network of three competing
species. The tree (top) tends not to co-occur with either of the two
shrub species, as indicated by the strongly negative coefficient linking
them. The two shrub species also compete with one another, but more
weakly (circled coefficient). \textbf{B.} In spite of the competitive
interactions between the two shrub species, their shared tendency to
occur in locations without trees makes their occurrence vectors
positively correlated (circled). \textbf{C.} Controlling for trees with
a conditional (all-else-equal) approach such as a partial covariance or
a Markov network leads to correct identification of the negative
shrub-shrub interaction (circled).

\noindent \textbf{Figure 2: A.} A small Markov network, defined by its
\(\alpha\) and \(\beta\) values. The abiotic environment favors the
occurrence of each species (\(\alpha >0\)), particularly species 2
(\(\alpha_2 > \alpha_1\)). The negative \(\beta_{12}\) coefficient is
consistent with competition between the two species. \textbf{B.} The
coefficients determine the probabilities of all four possible
presence-absence combinations for Species 1 and Species 2. \(\alpha_1\)
is added to the exponent whenever Species 1 is present (\(y_1 = 1\)),
but not when it is absent (\(y_1 = 0\)). Similarly, the exponent
includes \(\alpha_2\) only when species \(2\) is present (\(y_2 = 1\)),
and includes \(\beta_{12}\) only when both are present (\(y_1y_2 = 1\)).
The normalizing constant \(Z\), ensures that the four probabilities sum
to 1. In this case, \(Z\) is about 18.5. \textbf{C.} The expected
frequencies of all possible co-occurrence patterns between the two
species of interest, as calculated in the previous panel. \textbf{D.}
Without competition (i.e.~with \(\beta_{12}=0\), each species would
occur more often.

\noindent \textbf{Figure 3:} Proportion of variance in interaction
coefficients explained by each method versus number of sampled locations
across the three simulation types.

\noindent \textbf{Figure 4: A.} The Markov network's estimated
interaction coefficients were generally very similar to the GLM
estimates. \textbf{B.} The null model's estimates typically matched the
(negative) correlation coefficient, after controlling for landscape
size. \textbf{C.} For any given interaction strength, the null model was
much more likely to misclassify its sign with 95\% confidence than the
Markov network was.

@@

>>>>

\begin{spacing}{1.9}
\begin{flushleft}
\chead{\ifthenelse{\value{page}=1}{}{\textit{Inferring species interactions}}}
\renewcommand{\headrulewidth}{0pt}

@@

I then generated landscapes with up to 20 interacting species at 25,
200, or 1600 sites using three increasingly complex models (50
replicates for each combination of size and model). As described in
Appendix 2, I randomly drew the ``true'' coefficient values for each
replicate landscape so that most species pairs interacted negligibly, a
few pairs interacted very strongly, and competition was three times more
common than facilitation.

The first set of 20-species landscapes, like the landscapes with three
species, were generated directly from a Markov network to ensure that
the model could recover the parameters used to generate the ``observed''
co-occurrence data. Then, I added two environmental factors that varied
from location to location across the simulated landscapes, and simulated
a new set of co-occurrence data so that species' \(\alpha\) coefficients
depended on the local environment. The latter set of simulated
landscapes provide an important test of the methods' ability to
distinguish co-occurrence patterns that were generated from pairwise
biotic interactions from those that were generated by external forces
like abiotic environmental filtering. This task was made especially
difficult because---as with most analyses of presence-absence data for
co-occurrence patterns---the inference procedure did not have access to
any information about the environmental or spatial variables that helped
shape the landscape (cf Connor et al. 2013). I generated the final set
of landscapes by simulating the population dynamics of 20 species with
given per-capita interaction rates, as described in Appendix 3.

\noindent \textbf{Recovering species interactions from simulated data.}
I compared seven techniques for determining the sign and strength of the
associations between pairs of species from simulated data (Appendix 4).
First, I used the rosalia package (Harris 2015a) to fit Markov network
models, as described above. For the analyses with 20 species, I added a
weak regularizer (equivalent to a logistic prior with scale 2) to ensure
that the model estimates were always finite.

@@

\noindent
\textbf{Twenty species.} In general, each model's performance was
highest for large landscapes with simple assembly rules and no
environmental heterogeneity (Figure 3). Despite some variability across
contexts, the rank ordering across methods was very consistent. In
particular, the four methods that controlled for indirect effects (the
Markov network, the generalized linear models, and the two partial
correlation-based methods) always matched or outperformed those that did
not.\\Of these, the Markov network consistently performed the best. As
anticipated by Lee and Hastie (2012), generalized linear models closely
approximated the Markov network estimates, especially when the data sets
were very large. As reviewed in Gotelli and Ulrich (2009), however, most
analyses in this field involve fewer than 50 sites, so model performance
on the smaller landscapes where the gap was larger is most important. As
shown in Appendix 5, the standard errors associated with the estimates
in Figure 3 are small (less than 0.01), so the differences among methods
should not be attributed to sampling error.

@@

Finally, we can evaluate the models' statistical inferences (focusing on
the first two simulation types, for which the true interaction rates are
easiest to interpret). The Markov network's approximate Type I error
rate (defined here as the probability that 0 fell outside the 95\%
confidence interval for a pair of species where \(|\beta_{ij}|<0.1\))
depended on the simulation type: 0.03 for simulations that matched the
model's assumptions, versus 0.15 for simulations that included
environmental heterogeneity (Appendix 5). Coverage was similar for other
values of \(\beta\). For the null model, these rates were 0.31 and 0.50,
respectively---far higher than the nominal 0.05 rate. Figure 4C shows,
across a range of true interaction strengths, the probability that the
null model or the Markov network will predict the wrong sign of the
interaction with 95\% confidence. The null model makes such errors
nearly 8 times as often as the Markov network, even though it only
reject the null hypothesis twice as often overall (Appendix 5). The
Markov network's errors were also more concentrated around 0, so it
never misclassified strong interactions like the null model did.

@@

@@

\setlength{\parindent}{0cm} \setlength{\leftskip}{0cm}

\textbf{Figure 1: A.} A small network of three competing species. The
tree (top) tends not to co-occur with either of the two shrub species,
as indicated by the strongly negative coefficient linking them. The two
shrub species also compete with one another, but more weakly (circled
coefficient). \textbf{B.} In spite of the competitive interactions
between the two shrub species, their shared tendency to occur in
locations without trees makes their occurrence vectors positively
correlated (circled). \textbf{C.} Controlling for trees with a
conditional (all-else-equal) approach such as a partial covariance or a
Markov network leads to correct identification of the negative
shrub-shrub interaction (circled).

\textbf{Figure 2: A.} A small Markov network, defined by its \(\alpha\)
and \(\beta\) values. The abiotic environment favors the occurrence of
each species (\(\alpha >0\)), particularly species 2
(\(\alpha_2 > \alpha_1\)). The negative \(\beta_{12}\) coefficient is
consistent with competition between the two species. \textbf{B.} The
coefficients determine the probabilities of all four possible
presence-absence combinations for Species 1 and Species 2. \(\alpha_1\)
is added to the exponent whenever Species 1 is present (\(y_1 = 1\)),
but not when it is absent (\(y_1 = 0\)). Similarly, the exponent
includes \(\alpha_2\) only when species \(2\) is present (\(y_2 = 1\)),
and includes \(\beta_{12}\) only when both are present (\(y_1y_2 = 1\)).
The normalizing constant \(Z\), ensures that the four probabilities sum
to 1. In this case, \(Z\) is about 18.5. \textbf{C.} The expected
frequencies of all possible co-occurrence patterns between the two
species of interest, as calculated in the previous panel. \textbf{D.}
Without competition (i.e.~with \(\beta_{12}=0\), each species would
occur more often.

\textbf{Figure 3:} Proportion of variance in interaction coefficients
explained by each method versus number of sampled locations across the
three simulation types.

\textbf{Figure 4: A.} The Markov network's estimated interaction
coefficients were generally very similar to the GLM estimates.
\textbf{B.} The null model's estimates typically matched the (negative)
correlation coefficient, after controlling for landscape size.
\textbf{C.} For any given interaction strength, the null model was much
more likely to misclassify its sign with 95\% confidence than the Markov
network was.

### 2015-11-05 Harris_2015/simulation.Rmd
add fakedata and updates

The following function draws random coefficients for a Markov network of a pre-specified size.
The probability distribution was chosen so that most species pairs interact weakly, a few
act strongly, and that most of the interactions were negative rather than positive.

```{r, cache=TRUE}
make_coefficients = function(n_spp){
  # Exponential distribution has lots of mass near 0 but has
  # a long tail.
  true_beta_magnitudes = rexp(choose(n_spp, 2))

  # Species interactions are negative 3/4 of the time in 
  # these simulations.
  b = true_beta_magnitudes * sample(
    c(-1, 1), 
    size = length(true_beta_magnitudes), 
    prob = c(.75, .25),
    replace = TRUE
  )

  # Species' intercepts are normally distributed
  a = rnorm(n_spp, -1)

@@

I used the following function to generate "true" parameters with the method above, 
simulate a presence-absence landscape based on those parameters using Markov chain Monte Carlo, and save the results to a "fakedata" folder.

@@

```{r, cache = TRUE}
simulate_data = function(n_spp, n_sites, rep_name, n_gibbs = 1000, n_env = 2, sd = 2){

  # Determine the "true" parameters for the simulated assemblage
  par = make_coefficients(n_spp)

@@

  # Landscape starts as if betas were all zero. Each species' occurrence probability
  # depends on its alpha value and on the environment (assuming alpha_env is not zero).
  x = matrix(
    plogis(rep(1, n_sites) %*% t(alpha) + env %*% alpha_env), 
    nrow = n_sites, 
    ncol = n_spp
  )

@@

    # Each round of Gibbs sampling updates one species (column) across all sites
    # according to its conditional probability (i.e. conditional on environment
    # and the other species that are present).
    for(j in 1:n_spp){
      x[,j] = rbinom(
        nrow(x),
        size = 1,
        prob = plogis(x %*% beta[ , j] + alpha[j] + env %*% alpha_env[,j])
      )
    }
  }

@@

  file_stem = paste(n_spp, n_sites, rep_name, sep = "-")

@@

@@

for(n_sites in c(25, 200, 1600)){
  for(type in c("env", "no_env")){
    for(i in 1:50){
      simulate_data(
        n_spp = n_spp, 
        n_sites = n_sites, 
        rep_name = paste0(type, i),
        sd = ifelse(type == "env", 2, 0)
      )
    }
  }
}

>>>>

The following function draws random coefficients for a Markov network of a pre-specified size.

```{r, cache=TRUE}
make_coefficients = function(n_spp, p_neg, mean_alpha){
  # Exponential distribution has lots of mass near 0 but has
  # a long tail.
  true_beta_magnitudes = rexp(choose(n_spp, 2))

  # Multiply some proportion of the interactions
  # by -1
  b = true_beta_magnitudes * sample(
    c(-1, 1), 
    size = length(true_beta_magnitudes), 
    prob = c(p_neg, 1 - p_neg),
    replace = TRUE
  )

  # Species' intercepts are normally distributed
  a = rnorm(n_spp, mean_alpha)

@@

I used the following R function to generate "true" parameters with the method above, 
simulate a presence-absence landscape based on those parameters using Markov chain Monte Carlo, and save the results to a "fakedata" folder.

@@

For the abundance-based simulations, abundance was simulated with a Poisson
distribution with rate parameter given by 
$\mathrm{log}(1 + exp(\alpha_i + \sum_j\beta_{ij}y_j))$.

```{r, cache = TRUE}
simulate_data = function(n_spp, n_sites, rep_name, n_gibbs, n_env, sd, f, rdist, p_neg, mean_alpha){
  # n_spp:     number of species to include in the landscape
  # n_sites:   number of sites to include in the landscape
  # rep_name:  an identifier to use for the landscape replicate
  # n_gibbs:   number of Gibbs samples to perform
  # n_env:     number of environmental variables to simulate
  # sd:        standard deviation of environmental variables (can be zero)
  # f:         inverse link function (see above for two examples)
  # rdist:     a function for sampling a random value from a distribution
  # p_neg:     proportion of negative interactions (e.g. competition)
  # mean_alpha: the intercept value for the average species

  # Determine the "true" parameters for the simulated assemblage
  par = make_coefficients(n_spp, p_neg, mean_alpha)

@@

  # Landscape starts as if betas were all zero. Each species' occurrence probability
  # or abundance depends on its alpha value and on the environment (assuming alpha_env 
  # is not zero).
  x = matrix(
    f(rep(1, n_sites) %*% t(alpha) + env %*% alpha_env), 
    nrow = n_sites, 
    ncol = n_spp
  )

@@

    # Each round of Gibbs sampling updates one species (column) across all sites
    # according to its conditional probability (i.e. conditional on environment
    # and the other species that are present).
    for(j in 1:n_spp){
      x[,j] = rdist(
        nrow(x),
        f(x %*% beta[ , j] + alpha[j] + env %*% alpha_env[,j])
      )
    }
  }

  # Collapse abundance data to presence/absence and store
  # it as integer values rather than true/false
  x = x > 0
  mode(x) = "integer"

  colnames(x) = paste0("V", 1:n_spp)

@@

  file_stem = paste(n_sites, rep_name, sep = "-")

@@

```{r}
# Define a convenience function for Bernoulli random samples
rbern = function(n, prob){
  rbinom(n = n, size = 1, prob = prob)
}
```

@@

for(n_sites in c(25, 200, 1600)){
  for(type in c("no_env", "env", "abund")){
    for(i in 1:50){
      simulate_data(
        n_spp = n_spp, 
        n_sites = n_sites, 
        n_gibbs = 1000,
        n_env = 2,
        rep_name = paste0(type, i),
        sd = ifelse(type == "env", 2, 0),
        f = if(type == "abund"){function(x){log(1 + exp(x))}}else{plogis},
        rdist = if(type == "abund"){rpois}else{rbern},
        p_neg = if(type == "abund"){1}else{0.75},
        mean_alpha = if(type == "abund"){-1}else{5}
      )
    }
  }
}
